Synthesizing Theories of Human Language with Bayesian Program Induction

### 2018-06-25 manuscript.tex
more introduction

An age-old aspiration within artificial intelligence research is to
build a machine that helps automate the scientific process by
synthesizing theories or
models~\cite{paul1990autonomous,langley1987scientific,schmidt2009distilling}.
This aspiration remains largely unrealized: despite a few
small-scale successes on toy problems~\cite{Langley1981BACON5TD},
practicing scientists do not use machines to generate theories.
In contrast, AI has made great strides
on problems like machine vision and natural language processing.  How
can the artificial intelligence community get theory induction off the
ground?  This is an especially difficult question, because the current
mainstream in machine intelligence focuses on qualitatively different
classes of problems (e.g., prediction tasks like classification and
regression), whereas theory induction requires synthesizing
human-understandable causal models of real-world
phenomena~\cite{pearl2009causality}, so that human scientists can
understand and learn from the AI's outputs.

We propose theory induction research start with theories of human
language, and introduce a model of theory induction for a
key module of natural language: \emph{morphophonology}, the
relationship between word pronunciation and meaning.
Acquiring the morphophonology of a language involve solving a basic
problem confronting both linguists and children: given a collection
of utterances, together with aspects of their meaning, what is the
causal relationship between form and meaning?
\begin{comment}
Morphophonology, and
natural language more broadly, presents a tractable challenge to
theory induction, for three reasons.  First, every natural language is
a different theory induction problem, so there is a wide spectrum of
induction tasks to evaluate on.  Second, linguistics as a field
traditionally formalizes theories in computational terms, and so there
is a pre-existing body of representations and algorithms for the AI
researcher to draw upon.  Third, we know that acquiring theories of
language (e.g., grammars) is tractable, because children 
acquire language, and field linguists can be trained to build theories of grammar.
\end{comment}

Our contribution is a model for synthesizing theories of natural
language morphophonology.  Like linguists, the model starts with a
collection of utterances paired with meanings, and then constructs a
causal, interpretable model explaining how those meanings gave rise to
the utterances, roughly mirroring the process by which theorists
construct theories starting from experimental data.  In addition to
building theories, scientists distill those theories into higher-level
kinds of knowledge spanning multiple theories (e.g., energy
conservation and Lorentz invariance in physics; XXX in chemistry;
universal grammar in linguistics).  We argue that this higher-level
abstract knowledge is a crucial part of theory induction, imparting
prior knowledge and constraints on what would otherwise be an
ill-posed inductive reasoning problem.  Our model also acquires this
higher-level knowledge by jointly inferring theories for multiple
languages, and then shifting its inductive bias over theories to more
closely match the attested distribution of languages.  We evaluate our
algorithm on \# data sets spanning \# languages, automatically finding
theories that can model a wide swath of a core component of human
language.

>>>>

An age-old aspiration within artificial intelligence research is to
build a machine that helps automate the scientific process by
synthesizing theories or
models~\cite{paul1990autonomous,langley1987scientific,schmidt2009distilling}.
This aspiration remains largely unrealized: despite small-scale
demonstrations of machine-assisted theory
induction~\cite{Langley1981BACON5TD,schmidt2009distilling}, practicing
scientists do not use machines to generate theories.  In contrast, AI
has made great strides on problems like machine vision and natural
language processing.  How can the artificial intelligence community
get theory induction off the ground?  This is an especially difficult
question, because the current mainstream in machine intelligence
focuses on qualitatively different classes of problems (e.g.,
prediction tasks like classification and regression), whereas theory
induction requires synthesizing human-understandable causal models of
real-world phenomena~\cite{pearl2009causality}, so that human
scientists can understand and learn from the AI's outputs.

Theory induction is a challenge both for the natural sciences and AI,
with connections to questions in cognitive development.
Children, like scientists,
build systems of laws and concepts
to explain the world around them,
building intuitive theories of kinship, biology, physics, number,
grammar, and other domains, motivating the `child as scientist' metaphor~\cite{childAsScientist}.
To bridge these different kinds of theory induction---in scientists, children, and machines---
we propose theory induction research start with theories of human language,
for several reasons.
First, scientists, specifically linguists,
%% build theories of language starting from
%% experimental data (utterances paired with their meanings),
%% and 
have compiled large corpora 
from a variety of languages,
giving a rich and varied dataset for benchmarking theory induction algorithms.
Second, children easily acquire language from
modest amounts of data,
suggesting that
inducing theories of language is tractable,
even from sparse data;
and also suggesting that accounts of
linguistic theory induction
could shed light on language acquisition.
Going back to at least Chomsky,
linguists have sometimes thought of the acquisition problem as
approximating the problem facing the linguist~\cite{chomsky1956three,chomsky1988current}.
Third,
theories of language
are traditionally formalized in computational terms,
exposing a suite of formalisms ready to be deployed by AI researchers.
These three features of human language --- the availability of many highly-varied datasets,
the interfaces with cognitive development,
and the computational formalisms within linguistics ---
conspire to single out language as an especially suitable target for
research in automated theory induction.

We have proposed theory induction research start with theories of human
language, and here introduce a model of theory induction for a
key module of natural language: \emph{morphophonology}, the
relationship between word pronunciation and meaning.
Acquiring the morphophonology of a language involve solving a basic
problem confronting both linguists and children: given a collection
of utterances, together with aspects of their meaning, what is the
causal relationship between form and meaning?
\begin{comment}
Morphophonology, and
natural language more broadly, presents a tractable challenge to
theory induction, for three reasons.  First, every natural language is
a different theory induction problem, so there is a wide spectrum of
induction tasks to evaluate on.  Second, linguistics as a field
traditionally formalizes theories in computational terms, and so there
is a pre-existing body of representations and algorithms for the AI
researcher to draw upon.  Third, we know that acquiring theories of
language (e.g., grammars) is tractable, because children 
acquire language, and field linguists can be trained to build theories of grammar.
\end{comment}
Our contribution is a model for synthesizing theories of natural
language morphophonology.  Like linguists, the model starts with a
collection of utterances paired with meanings, and then constructs a
causal, interpretable model explaining how those meanings gave rise to
the utterances, roughly mirroring the process by which theorists
construct theories starting from experimental data.  In addition to
building theories, scientists distill those theories into higher-level
kinds of knowledge spanning multiple theories (e.g., energy
conservation and Lorentz invariance in physics; XXX in chemistry;
universal grammar in linguistics).  We argue that this higher-level
abstract knowledge is a crucial part of theory induction, imparting
prior knowledge and constraints on what would otherwise be an
ill-posed inductive reasoning problem.  Our model also acquires this
higher-level knowledge by jointly inferring theories for multiple
languages, and then shifting its inductive bias over theories to more
closely match the attested distribution of languages.  We evaluate our
algorithm on \# data sets spanning \# languages, automatically finding
theories that can model a wide swath of a core component of human
language.
